Tracing the Meta-Level: PyPy's Tracing JIT Compiler

### 2009-04-07 talk/icooolps2009/paper.tex
small tweaks. can remove most of the compression stuff again, as there is an official compressing style sheet.

\documentclass{acm_proc_article-sp}

@@

\numberofauthors{3}
\author{
\alignauthor Carl Friedrich Bolz\\
       \affaddr{Heinrich-Heine-Universität Düsseldorf}\\
       \affaddr{STUPS Group}\\
       \affaddr{Germany}\\
       \email{cfbolz@gmx.de}
\alignauthor Antonio Cuni\\
       \affaddr{University of Genova}\\
       \affaddr{DISI}\\
       \affaddr{Italy}\\
       \email{cuni@disi.unige.it}
\alignauthor Armin Rigo\\
       \email{arigo@tunes.org}
\alignauthor Maciej Fijalkowski\\
       \email{fijal@merlinux.eu}
}
\maketitle

@@

The contributions of this paper are:
\vspace{-0.3cm}
\begin{zitemize}
\item Applying a tracing JIT compiler to an interpreter.
\item Finding techniques for improving the generated code.
\item Integrating
\end{zitemize}

@@

Tracing JITs are built on the following basic assumptions:
\vspace{-0.3cm}
\begin{zitemize}
 \item programs spend most of their runtime in loops
 \item several iterations of the same loop are likely to take similar code paths
\end{zitemize}

@@

Typically, programs executed by a tracing VMs goes through various phases:
\vspace{-0.3cm}
\begin{zitemize}
\item Interpretation/profiling
\item Tracing
\item Code generation
\item Execution of the generated code
\end{zitemize}

@@

This process assumes that the path through the loop that was traced is a
``typical'' example of possible paths. Of course
it is possible that later another path through the loop is taken, in which case
one of the guards that were put into the machine code will fail.\footnote{There are more
complex mechanisms in place to still produce more code for the cases of guard
failures \cite{XXX}, but they are independent of the issues discussed in this
paper.}

@@

Let's look at a small example. Take the following (slightly contrived) RPython
code:
{\small
\begin{verbatim}
def f(a, b):
    if b % 46 == 41:
        return a - b
    else:
        return a + b
def strange_sum(n):
    result = 0
    while n >= 0:
        result = f(result, n)
        n -= 1
    return result
\end{verbatim}
}
\vspace{-0.4cm}
To trace this, a bytecode form of these functions needs to be introduced that
the tracer understands. The tracer interprets a bytecode that is an encoding of
the intermediate representation of PyPy's translation toolchain after type
inference has been performed.
When the profiler shows
that the \texttt{while} loop in \texttt{strange\_sum} is executed often the
tracing JIT will start to trace the execution of that loop.  The trace would
look as follows:
{\small
\begin{verbatim}
loop_header(result0, n0)
i0 = int_mod(n0, Const(46))
i1 = int_eq(i0, Const(41))
guard_false(i1)
result1 = int_add(result0, n0)
n1 = int_sub(n0, Const(1))
i2 = int_ge(n1, Const(0))
guard_true(i2)
jump(result1, n1)
\end{verbatim}
}
\vspace{-0.4cm}
The operations in this sequence are operations of the above-mentioned intermediate
representation (e.g. note that the generic modulo and equality operations in the
function above have been recognized to always take integers as arguments and are thus
rendered as \texttt{int\_mod} and \texttt{int\_eq}). The trace contains all the
operations that were executed in SSA-form \cite{cytron_efficiently_1991} and ends with a jump
to its beginning, forming an endless loop that can only be left via a guard
failure. The call to \texttt{f} is inlined into the trace. Note that the trace
contains only the hot \texttt{else} case of the \texttt{if} test in \texttt{f},
while the other branch is implemented via a guard failure. This trace can then
be converted into machine code and executed.

@@

A tracing JIT compiler finds the hot loops of the program it is compiling. In
our case, this program is the language interpreter. The most important hot interpreter loop
is the bytecode dispatch loop (for many simple
interpreters it is also the only hot loop).  Tracing one iteration of this
loop means that
the recorded trace corresponds to execution of one opcode. This means that the
assumption made by the tracing JIT -- that several iterations of a hot loop
take the same or similar code paths -- is wrong in this case. It is very
unlikely that the same particular opcode is executed many times in a row.
\begin{figure}
\input{code/tlr-paper.py}
\vspace{-0.4cm}
\caption{A very simple bytecode interpreter with registers and an accumulator.}
\label{fig:tlr-basic}
\end{figure}
\begin{figure}
{\small
\begin{verbatim}
    MOV_A_R     0   # i = a
    MOV_A_R     1   # copy of 'a'

@@

    MOV_R_A     2   # return res
    RETURN_A
\end{verbatim}
}
\vspace{-0.4cm}
\caption{Example bytecode: Compute the square of the accumulator}
\label{fig:square}
\end{figure}

@@

\begin{figure}
\input{code/normal-tracing.txt}
\vspace{-0.4cm}
\caption{Trace when executing the \texttt{DECR\_A} opcode}
\label{fig:trace-normal}
\end{figure}

@@

\begin{figure}
\input{code/tlr-paper-full.py}
\vspace{-0.4cm}
\caption{Simple bytecode interpreter with hints applied}
\label{fig:tlr-full}
\end{figure}

@@

For the small example the hints look like a lot of work. However, the number of
hints is essentially constant no matter how large the interpreter is, which
makes it seem less significant for larger interpreters.

@@

\begin{figure}
\input{code/no-green-folding.txt}
\vspace{-0.4cm}
\caption{Trace when executing the Square function of Figure \ref{fig:square},
with the corresponding bytecodes as comments.}
\label{fig:trace-no-green-folding}
\end{figure}

@@

\begin{figure}
\input{code/full.txt}
\vspace{-0.4cm}
\caption{Trace when executing the Square function of Figure \ref{fig:square},
with the corresponding opcodes as comments. The constant-folding of operations
on green variables is enabled.}
\label{fig:trace-full}
\end{figure}

>>>>

%\documentclass{acm_proc_article-sp}
\documentclass{sig-alternate}

@@

\numberofauthors{3}
\author{
\alignauthor Carl Friedrich Bolz\\
       \affaddr{Heinrich-Heine-Universität Düsseldorf}\\
       \affaddr{STUPS Group}\\
       \affaddr{Germany}\\
       \email{cfbolz@gmx.de}
\alignauthor Antonio Cuni\\
       \affaddr{University of Genova}\\
       \affaddr{DISI}\\
       \affaddr{Italy}\\
       \email{cuni@disi.unige.it}
\and
\alignauthor Armin Rigo\\
       \email{arigo@tunes.org}
\alignauthor Maciej Fijalkowski\\
       \email{fijal@merlinux.eu}
}
\maketitle

@@

The contributions of this paper are:
\begin{zitemize}
\item Applying a tracing JIT compiler to an interpreter.
\item Finding techniques for improving the generated code.
\end{zitemize}

@@

Tracing JITs are built on the following basic assumptions:
\begin{zitemize}
 \item programs spend most of their runtime in loops
 \item several iterations of the same loop are likely to take similar code paths
\end{zitemize}

@@

Typically, programs executed by a tracing VMs goes through various phases:
\begin{zitemize}
\item Interpretation/profiling
\item Tracing
\item Code generation
\item Execution of the generated code
\end{zitemize}

@@

This process assumes that the path through the loop that was traced is a
``typical'' example of possible paths. Of course
it is possible that later another path through the loop is taken, in which case
one of the guards that were put into the machine code will fail.\footnote{There are more
complex mechanisms in place to still produce more code for the cases of guard
failures \cite{andreas_gal_incremental_2006}, but they are independent of the issues discussed in this
paper.}

@@

Let's look at a small example. Take the following (slightly contrived) RPython
code:
{\small
\begin{verbatim}
def f(a, b):
    if b % 46 == 41:
        return a - b
    else:
        return a + b
def strange_sum(n):
    result = 0
    while n >= 0:
        result = f(result, n)
        n -= 1
    return result
\end{verbatim}
}
To trace this, a bytecode form of these functions needs to be introduced that
the tracer understands. The tracer interprets a bytecode that is an encoding of
the intermediate representation of PyPy's translation toolchain after type
inference has been performed.
When the profiler shows
that the \texttt{while} loop in \texttt{strange\_sum} is executed often the
tracing JIT will start to trace the execution of that loop.  The trace would
look as follows:
{\small
\begin{verbatim}
loop_header(result0, n0)
i0 = int_mod(n0, Const(46))
i1 = int_eq(i0, Const(41))
guard_false(i1)
result1 = int_add(result0, n0)
n1 = int_sub(n0, Const(1))
i2 = int_ge(n1, Const(0))
guard_true(i2)
jump(result1, n1)
\end{verbatim}
}
The operations in this sequence are operations of the above-mentioned intermediate
representation (e.g. note that the generic modulo and equality operations in the
function above have been recognized to always take integers as arguments and are thus
rendered as \texttt{int\_mod} and \texttt{int\_eq}). The trace contains all the
operations that were executed in SSA-form \cite{cytron_efficiently_1991} and ends with a jump
to its beginning, forming an endless loop that can only be left via a guard
failure. The call to \texttt{f} is inlined into the trace. Note that the trace
contains only the hot \texttt{else} case of the \texttt{if} test in \texttt{f},
while the other branch is implemented via a guard failure. This trace can then
be converted into machine code and executed.

@@

A tracing JIT compiler finds the hot loops of the program it is compiling. In
our case, this program is the language interpreter. The most important hot interpreter loop
is the bytecode dispatch loop (for many simple
interpreters it is also the only hot loop).  Tracing one iteration of this
loop means that
the recorded trace corresponds to execution of one opcode. This means that the
assumption made by the tracing JIT -- that several iterations of a hot loop
take the same or similar code paths -- is wrong in this case. It is very
unlikely that the same particular opcode is executed many times in a row.
\begin{figure}
\input{code/tlr-paper.py}
\caption{A very simple bytecode interpreter with registers and an accumulator.}
\label{fig:tlr-basic}
\end{figure}
\begin{figure}
{\small
\begin{verbatim}
    MOV_A_R     0   # i = a
    MOV_A_R     1   # copy of 'a'

@@

    MOV_R_A     2   # return res
    RETURN_A
\end{verbatim}
}
\caption{Example bytecode: Compute the square of the accumulator}
\label{fig:square}
\end{figure}

@@

\begin{figure}
\input{code/normal-tracing.txt}
\caption{Trace when executing the \texttt{DECR\_A} opcode}
\label{fig:trace-normal}
\end{figure}

@@

\begin{figure}
\input{code/tlr-paper-full.py}
\caption{Simple bytecode interpreter with hints applied}
\label{fig:tlr-full}
\end{figure}

@@

For the small example the hints look like a lot of work. However, the number of
hints is essentially constant no matter how large the interpreter is, which
makes it less significant for larger interpreters.

@@

\begin{figure}
\input{code/no-green-folding.txt}
\caption{Trace when executing the Square function of Figure \ref{fig:square},
with the corresponding bytecodes as comments.}
\label{fig:trace-no-green-folding}
\end{figure}

@@

\begin{figure}
\input{code/full.txt}
\caption{Trace when executing the Square function of Figure \ref{fig:square},
with the corresponding opcodes as comments. The constant-folding of operations
on green variables is enabled.}
\label{fig:trace-full}
\end{figure}
